Typed Clojure: A Practical Optional Type System for Clojure

### 2012-09-12 dissertation/design-choices.tex
Added some type rules and type metafunctions

$$
\begin{tdisplay}{Syntax of Terms}
\begin{altgrammar}
  \exp{} &::=& (\exp{} \overrightarrow{\exp{}})         % function application, multiple arguments
             \alt \{ \overrightarrow{\exp{}\ \exp{}} \} % map literal
             &\mbox{Expressions} \\ 
   \v{} &::=& \k{} \alt \nil{}
               \alt \{ \overrightarrow{\v{}\ \v{}} \}   % map literal
              &\mbox{Values}
\end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Evaluation Contexts}
  \begin{altgrammar}
    \E{} &::=& % application rules
              (\E{}\ \overrightarrow{\exp{}})  % target first
              \alt (\v{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % arguments next
              % map rules
              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
              % vector rule
              \alt [ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}} ] &\mbox{Evaluation Contexts}
  \end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Operational Semantics}
\begin{array}{c}
{\inferrule[E-Assoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
  { {\Gamma \vdash (assoc\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
\\
{\inferrule[E-Dissoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
  {\Gamma \vdash (dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
\\
{\inferrule[E-GetMapExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_1\ has\ entry\ with\ key\ v_2} \\
    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1} }
  { {\Gamma \vdash (get\ v_1\ v_2) \hookrightarrow v_3} }} \\
\\
{\inferrule[E-GetMapNotExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_1\ has\ no\ entry\ with\ key\ v_2} \\
    {v_3 = \nil{}} }
  { {\Gamma \vdash (get\ v_1\ v_2) \hookrightarrow v_3} }}
\end{array} 
\end{tdisplay} 
$$

$$
\begin{tdisplay}{Type Syntax}
\begin{altgrammar}
  \T{} &::=& \Top \alt \Nil \alt \k{}
\end{altgrammar}
\end{tdisplay}{Type Syntax}
$$

>>>>

$$
\begin{tdisplay}{Syntax of Terms}
\begin{altgrammar}
  \exp{} &::=& (\c{} \overrightarrow{\exp{}})         % function application, multiple arguments
             \alt \{ \overrightarrow{\exp{}\ \exp{}} \} % map literal
             &\mbox{Expressions} \\ 
  \v{} &::=& \k{} \alt \nil{}
              \alt \{ \overrightarrow{\v{}\ \v{}} \}   % map literal
              &\mbox{Values} \\
  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
              &\mbox{Constants}
\end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Evaluation Contexts}
  \begin{altgrammar}
    \E{} &::=& % application rules
              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
              % map rules
              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
              &\mbox{Evaluation Contexts}
  \end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Operational Semantics}
\begin{array}{c}
{\inferrule[E-Assoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
    {{(\assoc{}\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
\\
{\inferrule[E-Dissoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
  {(dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
\\
{\inferrule[E-GetMapExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_1\ has\ entry\ with\ key\ v_2} \\
    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1} }
  { {(get\ v_1\ v_2) \hookrightarrow v_3} }} \\
\\
{\inferrule[E-GetMapNotExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_1\ has\ no\ entry\ with\ key\ v_2} \\
    {v_3 = \nil{}} }
  { {(get\ v_1\ v_2) \hookrightarrow v_3} }}
\end{array} 
\end{tdisplay} 
$$

$$
\begin{tdisplay}{Type Syntax}
\begin{altgrammar}
  \T{} &::=& \Top \alt \Bottom \alt \Nil \alt \k{} \alt \{ \overrightarrow{\k{}\ \T{}} \}
             \alt {\IPersistentMap {\T{}} {\T{}}} \alt ( \cup\ \overrightarrow{\T{}} )
             \alt ( \cap\ \overrightarrow{\T{}} )
\end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Core Type Rules}
\begin{array}{c}
{\inferrule[T-AssocHMap]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\k{}}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
    {\vdash {\issubtype {{\T{}}_{m}} {{ \{ \} }} }} \\
    {\T{} = (update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})} }
  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
\\
{\inferrule[T-AssocHMapPromote]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
    {\vdash {\issubtype {{\T{}}_{m}} { \{ \}}}} \\
    {\T{} = (promote\_hmap\ {\T{}}_{m}\ {\T{}}_{k}\ {\T{}}_{v})} }
  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
\\
{\inferrule[T-AssocIPersistentMap]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}} }} \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}} }} \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}} }} }
  { {\Gamma \vdash {\hastype {(\assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {T}}}}}
\end{array}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Type Metafunctions}
\begingroup
\fontsize{10pt}{12pt}
\begin{altgrammar}
  (update\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\k{}}_1\ {\T{}}_1) = 
                 \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}}\ {\k{}}_1\ {\T{}}_1 \}\ \\
  (update\_hmap\ (\cap\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v} = 
                               (\cap\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})} \\
  (update\_hmap\ (\cup\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v} = 
                               (\cup\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
  \\ \\
  (promote\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\T{}}_{kn}\ {\T{}}_{vn})
                                    = {\IPersistentMap {(\cup\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
                                                       {(\cup\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
  (promote\_hmap\ {\IPersistentMap {{\T{}}_{k}} {{\T{}}_{v}}} \ {\T{}}_{kn}\ {\T{}}_{vn})
                                    = {\IPersistentMap {(\cup\ {\T{}}_{k}\ {\T{}}_{kn})}
                                                       {(\cup\ {\T{}}_{v}\ {\T{}}_{vn})}}
\end{altgrammar}
\endgroup
\end{tdisplay}
$$
